---
title: 'Advent Of Code 2021 -- Day02: Dive!'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Day 2: Dive!

# Puzzle 1

Now, you need to figure out how to pilot this thing.

It seems like the submarine can take a series of commands like forward 1, down 2, or up 3:

- *forward X* increases the horizontal position by X units.
- *down X* increases the depth by X units.
- *up X* decreases the depth by X units.

Note that since you're on a submarine, down and up affect your depth, and so they *have the opposite result* of what you might expect.

The submarine seems to already have a planned course (your puzzle input). You should probably figure out where it's going. For example:

`forward 5`
`down 5`
`forward 8`
`up 3`
`down 8`
`forward 2`

Your horizontal position and depth both start at 0. The steps above would then modify them as follows:

forward 5 adds 5 to your horizontal position, a total of 5.
down 5 adds 5 to your depth, resulting in a value of 5.
forward 8 adds 8 to your horizontal position, a total of 13.
up 3 decreases your depth by 3, resulting in a value of 2.
down 8 adds 8 to your depth, resulting in a value of 10.
forward 2 adds 2 to your horizontal position, a total of 15.

After following these instructions, you would have a horizontal position of `15` and a depth of `10.` (Multiplying these together produces `150`.)

Calculate the horizontal position and depth you would have after following the planned course. What do you get if you multiply your final horizontal position by your final depth?


```{r input}
input <- readLines("input_day02.txt")
```


```{r puzzle1}
puz1 <- 
  tibble(input = input) %>% 
    # could do this all in one shot, but more easily understandable as different evals
  mutate(type = if_else(str_detect(input, "forward"), "horiz", "depth"), 
         value = parse_number(input), 
         posneg = if_else(str_detect(input, "up"), -1, 1), 
         final_val = value * posneg) 

puz1 %>% 
  count(type, wt = final_val, name = "total") %>% 
  pull(total) %>% 
  prod()

# count function above works the same as below:
# group_by(type) %>% 
# summarize(total = sum(final_val))
```

Multiplying the final horizontal position (1895) by the final depth (894) we get a result of `1694130`



# Puzzle 2

Based on your calculations, the planned course doesn't seem to make any sense. You find the submarine manual and discover that the process is actually slightly more complicated.

In addition to horizontal position and depth, you'll also need to track a third value, aim, which also starts at 0. The commands also mean something entirely different than you first thought:

- *down X* increases your aim by X units.
- *up X* decreases your aim by X units.
- *forward X* does two things:
    It increases your horizontal position by X units.
    It increases your depth by your aim multiplied by X.

Again note that since you're on a submarine, down and up do the opposite of what you might expect: "down" means aiming in the positive direction.

Now, the above example does something different:

forward 5 adds 5 to your horizontal position, a total of 5. Because your aim is 0, your depth does not change.
down 5 adds 5 to your aim, resulting in a value of 5.
forward 8 adds 8 to your horizontal position, a total of 13. Because your aim is 5, your depth increases by 8*5=40.
up 3 decreases your aim by 3, resulting in a value of 2.
down 8 adds 8 to your aim, resulting in a value of 10.
forward 2 adds 2 to your horizontal position, a total of 15. Because your aim is 10, your depth increases by 2*10=20 to a total of 60.

After following these new instructions, you would have a horizontal position of `15` and a depth of `60` (Multiplying these produces 900.)

Using this new interpretation of the commands, calculate the horizontal position and depth you would have after following the planned course. What do you get if you multiply your final horizontal position by your final depth?


```{r puzzle2}
# initial settings
puz2 <- c(horiz = 0L, depth = 0L, aim = 0L)

# functions for each instruction
down <- function (value, .vec = puz2) { 
  .vec["aim"] <- .vec["aim"] + value 
  .vec
}

up <- function(value, .vec = puz2) { 
  .vec["aim"] <- .vec["aim"] - value 
  .vec
}

forward <- function(value, .vec = puz2) {
  .vec["horiz"] <- .vec["horiz"] + value
  .vec["depth"] <- .vec["depth"] + (.vec["aim"] * value)
  .vec
}

# for loop to iterate
for (i in seq_along(input)) {
  
    # split out instruction and value from input
  x <- str_split(input[i], " ")[[1]]
  instr <- x[1]
  val <- parse_integer(x[2])
  
  # if else to do correct function at each step
  if (instr == "down") {
    puz2 <- down(val)
  } else if (instr == "up") {
    puz2 <- up(val)
  } else if (instr == "forward") {
    puz2 <- forward(val)
  }
  
}
-
# finally, multiply our horizontal position by depth for answer
puz2["horiz"] * puz2["depth"]

```

Following these new instructions, we get a final answer of `1698850445` multiplying our final horizontal position (1895) by our final depth (896491)


Puzzle 2 without a loop:

```{r puzzle2 sans loop}
tibble(input = input) %>% 
  separate(input, into = c("cmd", "value"), sep = " ", convert = TRUE) %>% 
  mutate(value = if_else(cmd == "up", -value, value), 
         aim = cumsum(if_else(cmd != "forward", value, 0L)), 
         horiz = cumsum(if_else(cmd == "forward", value, 0L)), 
         depth = cumsum(if_else(cmd == "forward", aim * value, 0L)), 
         answer = horiz * depth) %>% 
   # pull last record as that's the last step
  tail(1) %>% 
  pull(answer)
  
```

Above also gets correct answer of `1698850445`



Puzzle 2 using `reduce` to iterate...

```{r puzzle 2 with reduce}
input
```

