---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 20)

library(tidyverse)
```

# Day 05: Hydrothermal Venture

## Puzzle 1

You come across a field of hydrothermal vents on the ocean floor! These vents constantly produce large, opaque clouds, so it would be best to avoid them if possible.

They tend to form in *lines*; the submarine helpfully produces a list of nearby lines of vents (your puzzle input) for you to review. For example:

`0,9 -> 5,9`
`8,0 -> 0,8`
`9,4 -> 3,4`
`2,2 -> 2,1`
`7,0 -> 7,4`
`6,4 -> 2,0`
`0,9 -> 2,9`
`3,4 -> 1,4`
`0,0 -> 8,8`
`5,5 -> 8,2`

Each line of vents is given as a line segment in the format x1,y1 -> x2,y2 where x1,y1 are the coordinates of one end the line segment and x2,y2 are the coordinates of the other end. These line segments include the points at both ends. In other words:

- An entry like 1,1 -> 1,3 covers points 1,1, 1,2, and 1,3.
- An entry like 9,7 -> 7,7 covers points 9,7, 8,7, and 7,7.

For now, *only consider horizontal and vertical lines*: lines where either x1 = x2 or y1 = y2.

So, the horizontal and vertical lines from the above list would produce the following diagram:

```
.......1..
..1....1..
..1....1..
.......1..
.112111211
..........
..........
..........
..........
222111....
```

In this diagram, the top left corner is 0,0 and the bottom right corner is 9,9. Each position is shown as *the number of lines which cover that point* or . if no line covers that point. The top-left pair of 1s, for example, comes from 2,2 -> 2,1; the very bottom row is formed by the overlapping lines 0,9 -> 5,9 and 0,9 -> 2,9.

To avoid the most dangerous areas, you need to determine *the number of points where at least two lines overlap*. In the above example, this is anywhere in the diagram with a 2 or larger - a total of 5 points.

Consider only horizontal and vertical lines. *At how many points do at least two lines overlap*?

```{r input}
test <- read_lines("0,9 -> 5,9
8,0 -> 0,8
9,4 -> 3,4
2,2 -> 2,1
7,0 -> 7,4
6,4 -> 2,0
0,9 -> 2,9
3,4 -> 1,4
0,0 -> 8,8
5,5 -> 8,2")

input <- read_lines("input_day05.txt")
```


This function will create:
(a) the table of evaluation from the input in the global environment. This is useful
as it has a column for each input on whether it is a straight line or not
(b) the list of matrices with each line segment marked with a 1 based on the evaluation
from table above

credit for `cbind` idea for subsetting matrices came from David Robinson on twitter
- https:://twitter.com/drob
- https://twitter.com/drob/status/1467367984192708609/photo/1

I was unaware, before seeing that tweet, that a matrix can be subset using another matrix
and that the `cbind` function will get us that matrix in order to subset. For vector
inputs, it essentially "combines" objects together element-wise and 
returns them in a matrix. 

For example...

```
cbind(1:5, 6:10)
     [,1] [,2]
[1,]    1    6
[2,]    2    7
[3,]    3    8
[4,]    4    9
[5,]    5   10

cbind(1:3, 4:6, 7:9)
     [,1] [,2] [,3]
[1,]    1    4    7
[2,]    2    5    8
[3,]    3    6    9
```

```{r parse input function}
parse_input <- function(.x) {
  
  mat_size <- 
    .x %>% 
    str_split(" -> ") %>% 
    map(.x = ., 
        .f = ~ str_split(., ",") %>% 
          unlist() %>% 
          max() ) %>% 
    unlist() %>% 
    parse_number() %>% 
    max() %>% 
    {. + 1} # add 1 because puzzle's coordinates index is at zero
  
  # create blank matrices
  .m <- vector("list", length = length(.x))
  
  .m <- map(.x = .m, 
            .f = ~ matrix(rep(0L, mat_size^2), nrow = mat_size))
  
  .t <- 
    tibble(input = .x) %>% 
    separate(input, into = c("start", "end"), sep = " -> ") %>% 
    separate(start, into = c("x0", "y0"), sep = ",") %>% 
    separate(end, into = c("x1", "y1"), sep = ",") %>% 
    mutate(across(.fns = ~ {parse_integer(.x) + 1L} ) ) %>% 
    # add 1 as indexing in puzzle starts at zero
    mutate(diffx = x1 - x0, 
           diffy = y1 - y0, 
           straight = if_else(diffy == 0L | diffx == 0L, TRUE, FALSE), 
           yeval = if_else(diffy == 0L, 
                           as.character(y1), 
                           str_c(as.character(y1), ":", as.character(y0))), 
           xeval = if_else(diffx == 0L, 
                           as.character(x1), 
                           str_c(as.character(x1), ":", as.character(x0))), 
           expr = str_c("cbind(", yeval, ", ", xeval, ")") 
           # cbind will let us create a combination of coordinates to subset a matrix to 
           #    mark as part of the line segment we care about
    )
  
  # create this table in global environment, as it has our eval on straight lines and such
  assign("table_eval", .t, envir = globalenv())
  
  
  for (i in seq_along(.m)) {
    
    # create full execution call as string, based on above evaluations
    # this will change to a 1 in each appropriate matrix where the line segment crosses
    .e <- str_c(".m[[", as.character(i), "]][", .t$expr[[i]],"] <- 1L" )
    
    # parse to make the character string created above into an "expression" type
    .e <- parse(text = .e)
    
    # evaluate and execute our expression
    eval(.e)
    
  }
  
  # create final matrices list object in global environment
  assign("matrices", .m, envir = globalenv())
}

```



For puzzle 1...
Consider only horizontal and vertical lines. *At how many points do at least two lines overlap*

```{r puzzle1}
parse_input(input)

puzzle_solve <- function(.tbl = table_eval, .mat = matrices, filter = TRUE) {
  
  # if filter is TRUE, then filter down to only straight lines
  if (filter) {
    
    # which inputs are straight lines?
    indices <- which(.tbl$straight == TRUE)
    
    # filter down to our matrices that only have straight lines
    puz <- .mat[indices]
    
  } else {
    
    puz <- .mat
    
  }
  
  # condense down our matrices using reduce
  # helpfully, adding (`+`) matrices is for each coordinate individually
  # we should get a single matrix counting the occurrence at each location
  puz <- reduce(puz, `+`)
  
  puz %>% 
    .[. >= 2] %>% 
    length()
}

puzzle_solve()

```

We have `4421` points where at least two straight lines overlap in our data.



## Puzzle 2

Unfortunately, considering only horizontal and vertical lines doesn't give you the full picture; you need to also consider diagonal lines.

Because of the limits of the hydrothermal vent mapping system, the lines in your list will only ever be horizontal, vertical, or a *diagonal line at exactly 45 degrees*. In other words:

- An entry like 1,1 -> 3,3 covers points 1,1, 2,2, and 3,3.
- An entry like 9,7 -> 7,9 covers points 9,7, 8,8, and 7,9.

Considering all lines from the above example would now produce the following diagram:

`1.1....11.`
`.111...2..`
`..2.1.111.`
`...1.2.2..`
`.112313211`
`...1.2....`
`..1...1...`
`.1.....1..`
`1.......1.`
`222111....`

You still need to determine *the number of points where at least two lines overlap*. In the above example, this is still anywhere in the diagram with a 2 or larger - now a total of 12 points.

Consider all of the lines. *At how many points do at least two lines overlap*?



Should be the same process as puzzle1, just without filtering for only straight lines...
Changed puzzle1 to be a function call instead so I could reuse code here for puzzle2

```{r puzzle2}
puzzle_solve(filter = FALSE)
```

Including diagonal lines, we get `18674` points where at least two lines overlap.