---
title: 'Advent Of Code 2021 -- Day02: Dive!'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

# Day 2: Dive!

# Puzzle 1

Now, you need to figure out how to pilot this thing.

It seems like the submarine can take a series of commands like forward 1, down 2, or up 3:

- *forward X* increases the horizontal position by X units.
- *down X* increases the depth by X units.
- *up X* decreases the depth by X units.

Note that since you're on a submarine, down and up affect your depth, and so they *have the opposite result* of what you might expect.

The submarine seems to already have a planned course (your puzzle input). You should probably figure out where it's going. For example:

`forward 5`
`down 5`
`forward 8`
`up 3`
`down 8`
`forward 2`

Your horizontal position and depth both start at 0. The steps above would then modify them as follows:

forward 5 adds 5 to your horizontal position, a total of 5.
down 5 adds 5 to your depth, resulting in a value of 5.
forward 8 adds 8 to your horizontal position, a total of 13.
up 3 decreases your depth by 3, resulting in a value of 2.
down 8 adds 8 to your depth, resulting in a value of 10.
forward 2 adds 2 to your horizontal position, a total of 15.

After following these instructions, you would have a horizontal position of `15` and a depth of `10.` (Multiplying these together produces `150`.)

Calculate the horizontal position and depth you would have after following the planned course. What do you get if you multiply your final horizontal position by your final depth?


```{r input}
input <- readLines("input_day02.txt")
```


```{r puzzle1}
puz1 <- 
  tibble(input = input) %>% 
  mutate(type = if_else(str_detect(input, "forward"), "horiz", "depth"), 
         value = parse_number(input), 
         posneg = if_else(str_detect(input, "up"), -1L, 1L), 
         final_val = value * posneg) %>% 
  count(type, wt = final_val, name = "total")
# count above functions same as below:
# group_by(type) %>% 
# summarize(total = sum(final_val))

prod(puz1$total)
```

Multiplying the final horizontal position (1895) by the final depth (894) we get a result of `1694130`



# Puzzle 2


```{r puzzle2}

```


