---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 20)

library(tidyverse)
```

# Day 04: Giant Squid

## Puzzle 1

You're already almost 1.5km (almost a mile) below the surface of the ocean, already so deep that you can't see any sunlight. What you *can* see, however, is a giant squid that has attached itself to the outside of your submarine.

Maybe it wants to play bingo?

Bingo is played on a set of boards each consisting of a 5x5 grid of numbers. Numbers are chosen at random, and the chosen number is *marked* on all boards on which it appears. (Numbers may not appear on all boards.) If all numbers in any row or any column of a board are marked, that board *wins*. (Diagonals don't count.)

The submarine has a *bingo subsystem* to help passengers (currently, you and the giant squid) pass the time. It automatically generates a random order in which to draw numbers and a random set of boards (your puzzle input). For example:

`7,4,9,5,11,17,23,2,0,14,21,24,10,16,13,6,15,25,12,22,18,20,8,19,3,26,1`
`              `
`22 13 17 11  0`
` 8  2 23  4 24`
`21  9 14 16  7`
` 6 10  3 18  5`
` 1 12 20 15 19`
`              `
` 3 15  0  2 22`
` 9 18 13 17  5`
`19  8  7 25 23`
`20 11 10 24  4`
`14 21 16 12  6`
`              `
`14 21 17 24  4`
`10 16 15  9 19`
`18  8 23 26 20`
`22 11 13  6  5`
` 2  0 12  3  7`

After the first five numbers are drawn (7, 4, 9, 5, and 11), there are no winners.

After the next six numbers are drawn (17, 23, 2, 0, 14, and 21), there are still no winners.

Finally, 24 is drawn.

At this point, the third board wins because it has at least one complete row or column of marked numbers (in this case, the entire top row is marked: `14 21 17 24 4`).

The *score* of the winning board can now be calculated. Start by finding the *sum of all unmarked numbers* on that board; in this case, the sum is 188. Then, multiply that sum by *the number that was just called* when the board won, 24, to get the final score, 188 * 24 = 4512.

To guarantee victory against the giant squid, figure out which board will win first. *What will your final score be if you choose that board?*

```{r input}
test <- "7,4,9,5,11,17,23,2,0,14,21,24,10,16,13,6,15,25,12,22,18,20,8,19,3,26,1

22 13 17 11  0
 8  2 23  4 24
21  9 14 16  7
 6 10  3 18  5
 1 12 20 15 19

 3 15  0  2 22
 9 18 13 17  5
19  8  7 25 23
20 11 10 24  4
14 21 16 12  6

14 21 17 24  4
10 16 15  9 19
18  8 23 26 20
22 11 13  6  5
 2  0 12  3  7"

input <- read_file("input_day04.txt")

# function to parse out the instructions as
#   (1) a character vector of instructions
#   (2) a list of matrices for the bingo cards
# Both are saved out to the global environment

parse_input <- function(.input) {
  
  .x <- str_split_fixed(.input, "\n\n", 2)
  
  .y <- 
    str_split(.x[, 1], ",") %>% 
    pluck(1) %>% 
    str_pad(., width = 2, side = "left", pad = "0")
  
  assign("instructions", .y, globalenv())
  
  .x <- .x[, -1]
  
  .x <- 
    str_split_fixed(.x, "\n\n", n = str_count(.x, "\n\n") + 1) %>% 
    as.list() %>% 
    map(.x = ., 
        .f = ~ str_replace_all(.x, "\n", " ") %>% 
          str_replace_all("\\s{1,}", ",") %>% 
          {str_split(., ",")[[1]]} %>% 
          .[. != ""] %>% 
          str_pad(., width = 2, side = "left", pad = "0") %>% 
          matrix(nrow = 5, byrow = TRUE) )
  
  assign("bingo", .x, globalenv())
}

# swap "test" or "input" here for which we want to run in puzzles
parse_input(input)

```


```{r puzzle1}
# str_replace returns a character vector and removes the "matrix" class
# using base gsub instead as it retains the "matrix" structure we want

# loop to process instructions
for (i in seq_along(instructions)) {
  
  v <- instructions[i]
  
  # mark with an "x" this bingo call
  bingo <- map(.x = bingo, 
               .f = ~ gsub(v, "x", .x))
  
  # rowsum check
  rsc <- 
    map(.x = bingo, 
        .f = ~ {. == "x"} %>% 
          rowSums() %>% 
          {. == 5} %>% 
          sum()
    ) %>% 
    unlist() %>% 
    sum()
  
  if (rsc > 0) {
    last_instr <- parse_number(v)
    fnc <- "rowSums"
    break
  }
  
  # colsum check
  csc <- 
    map(.x = bingo, 
        .f = ~ {. == "x"} %>% 
          colSums() %>% 
          {. == 5} %>% 
          sum()
    ) %>% 
    unlist() %>% 
    sum()
  
  if (csc > 0) {
    last_instr <- parse_number(v)
    fnc <- colSums
    break
  }
}


# now let's find which board is the one that caused the break
puz1_board_number <- 
  map(.x = bingo, 
      .f = ~ {. == "x"} %>% 
        rlang::exec(fnc, .) %>% 
        max()) %>% 
  unlist() %>% 
  {which(. == 5)}

puz1_board <- bingo[[puz1_board_number]]

# sum of all unmarked numbers by the last instructions processed in the answer to puzzle1
gsub("x", "0", puz1_board) %>% 
  parse_number() %>% 
  sum() %>% 
  {. * last_instr}
```

`12796` is the answer for our puzzle 1 input




## Puzzle 2


```{r puzzle2}

```

