---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 20)

library(tidyverse)
```

# Day 12: Passage Pathing 

## Puzzle 1

With your submarine's subterranean subsystems subsisting sub-optimally, the only way you're getting out of this cave anytime soon is by finding a path yourself. Not just a path - the only way to know if you've found the best path is to *find all of them*.

Fortunately, the sensors are still mostly working, and so you build a rough map of the remaining caves (your puzzle input). For example:

```
start-A
start-b
A-c
A-b
b-d
A-end
b-end
```

This is a list of how all of the caves are connected. You start in the cave named `start`, and your destination is the cave named `end`. An entry like b-d means that cave b is connected to cave d - that is, you can move between them.

So, the above cave system looks roughly like this:

```
    start
    /   \
c--A-----b--d
    \   /
     end
```

Your goal is to find the number of *distinct paths* that start at start, end at end, and don't visit small caves more than once. There are two types of caves: *big caves* (written in uppercase, like A) and *small caves* (written in lowercase, like b). It would be a waste of time to visit any small cave more than once, but big caves are large enough that it might be worth visiting them multiple times. So, *all paths you find should visit small caves at most once, and can visit big caves any number of times*.

Given these rules, there are 10 paths through this example cave system:

```
start,A,b,A,c,A,end
start,A,b,A,end
start,A,b,end
start,A,c,A,b,A,end
start,A,c,A,b,end
start,A,c,A,end
start,A,end
start,b,A,c,A,end
start,b,A,end
start,b,end
```

(Each line in the above list corresponds to a single path; the caves visited by that path are listed in the order they are visited and separated by commas.)

Note that in this cave system, cave d is never visited by any path: to do so, cave b would need to be visited twice (once on the way to cave d and a second time when returning from cave d), and since cave b is small, this is not allowed.

Here is a slightly larger example:

```
dc-end
HN-start
start-kj
dc-start
dc-HN
LN-dc
HN-end
kj-sa
kj-HN
kj-dc
```

The 19 paths through it are as follows:

```
start,HN,dc,HN,end
start,HN,dc,HN,kj,HN,end
start,HN,dc,end
start,HN,dc,kj,HN,end
start,HN,end
start,HN,kj,HN,dc,HN,end
start,HN,kj,HN,dc,end
start,HN,kj,HN,end
start,HN,kj,dc,HN,end
start,HN,kj,dc,end
start,dc,HN,end
start,dc,HN,kj,HN,end
start,dc,end
start,dc,kj,HN,end
start,kj,HN,dc,HN,end
start,kj,HN,dc,end
start,kj,HN,end
start,kj,dc,HN,end
start,kj,dc,end
```

Finally, this even larger example has 226 paths through it:

```
fs-end
he-DX
fs-he
start-DX
pj-DX
end-zg
zg-sl
zg-pj
pj-he
RW-he
fs-DX
pj-RW
zg-RW
start-pj
he-WI
zg-he
pj-fs
start-RW
```

*How many paths through this (your) cave system are there that visit small caves at most once?*

```{r input}
parse_input <- function(.x) {
  .t <- 
    tibble(input = read_lines(.x)) %>% 
    separate(input, into = c("i", "o"), sep = "-")
  
  bind_rows(.t, 
            filter(.t, i != "start", o != "end") %>% 
              rename(o = i, i = o) %>% 
              relocate(i))
  
}

test1 <- parse_input("start-A\nstart-b\nA-c\nA-b\nb-d\nA-end\nb-end\n")
test2 <- parse_input("dc-end\nHN-start\nstart-kj\ndc-start\ndc-HN\nLN-dc\nHN-end\nkj-sa\nkj-HN\nkj-dc\n")
test3 <- parse_input("fs-end\nhe-DX\nfs-he\nstart-DX\npj-DX\nend-zg\nzg-sl\nzg-pj\npj-he\nRW-he\nfs-DX\npj-RW\nzg-RW\nstart-pj\nhe-WI\nzg-he\npj-fs\nstart-RW\n")
input <- parse_input("input_day12.txt")
```


```{r puzzle1}
puz1 <- input

islower <- function(.x) {
  str_split(.x, "")[[1]] %>% 
    {all(. %in% letters)}
}

step_fun <- function(.x, .tbl = puz1) {
  # browser()
  .t <- tibble(v = .x) %>% 
    mutate(path = map(v, ~ str_split(.x, ",") %>% 
                        pluck(1)), 
           join = map_chr(path, ~ {.x[length(.x)]}))
  
  .t <- 
    .t %>% 
    rowwise() %>% 
    mutate(join = if_else(join %in% head(unlist(path), -1) & 
                            islower(join) & 
                            !(join %in% c("end", "fail") ), 
                          NA_character_, 
                          join)) %>% 
    ungroup()
  
  .t <- 
    .t %>% 
    left_join(.tbl, by = c("join" = "i")) %>% 
    mutate(o = case_when(join %in% c("end", "fail") ~ "", 
                         is.na(o) ~ "fail", 
                         TRUE ~ o), 
           new = if_else(o != "", str_c(v, o, sep = ","), v))
  
  pull(.t, new)
  
}
  
ends_fun <- function(.x = p1_string) {
  map_chr(.x, ~ str_split(.x, ",") %>% 
            pluck(1) %>% 
            {.[length(.)]}
          )
}

  
p1_string <- "start"
p1_ends <- ends_fun()

while ( !(all(p1_ends %in% c("end", "fail"))) ) {
  p1_string <- step_fun(p1_string)
  p1_ends <- ends_fun()
}
  

p1_string[p1_ends == "end"] %>% unique() %>% length()


```

Out input has `3802` unique paths that visit a specific small cave at most once in transit. 



## Puzzle 2


```{r puzzle2}

```

