---
title: ''
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(scipen = 20)

library(tidyverse)
```

# Day 14: Extended Polymerization

# Puzzle 1 

The incredible pressures at this depth are starting to put a strain on your submarine. The submarine has polymerization equipment that would produce suitable materials to reinforce the submarine, and the nearby volcanically-active caves should even have the necessary input elements in sufficient quantities.

The submarine manual contains instructions for finding the optimal polymer formula; specifically, it offers a *polymer template* and a list of *pair insertion rules* (your puzzle input). You just need to work out what polymer would result after repeating the pair insertion process a few times.

For example:

```
NNCB

CH -> B
HH -> N
CB -> H
NH -> C
HB -> C
HC -> B
HN -> C
NN -> C
BH -> H
NC -> B
NB -> B
BN -> B
BB -> N
BC -> B
CC -> N
CN -> C
```

The first line is the *polymer template* - this is the starting point of the process.

The following section defines the *pair insertion rules*. A rule like `AB -> C` means that when elements A and B are immediately adjacent, element C should be inserted between them. These insertions all happen simultaneously.

So, starting with the polymer template `NNCB`, the first step simultaneously considers all three pairs:

The first pair (NN) matches the rule `NN -> C`, so element C is inserted between the first N and the second N.
The second pair (NC) matches the rule `NC -> B`, so element B is inserted between the N and the C.
The third pair (CB) matches the rule `CB -> H`, so element H is inserted between the C and the B.

Note that these pairs overlap: the second element of one pair is the first element of the next pair. Also, because all pairs are considered simultaneously, inserted elements are not considered to be part of a pair until the next step.

After the first step of this process, the polymer becomes `NCNBCHB`.

Here are the results of a few steps using the above rules:

```
Template:     NNCB
After step 1: NCNBCHB
After step 2: NBCCNBBBCBHCB
After step 3: NBBBCNCCNBBNBNBBCHBHHBCHB
After step 4: NBBNBNBBCCNBCNCCNBBNBBNBBBNBBNBBCBHCBHHNHCBBCBHCB
```

This polymer grows quickly. After step 5, it has length 97; After step 10, it has length 3073. After step 10, B occurs 1749 times, C occurs 298 times, H occurs 161 times, and N occurs 865 times; taking the quantity of the most common element (B, 1749) and subtracting the quantity of the least common element (H, 161) produces `1749 - 161 = 1588`.

*Apply 10 steps of pair insertion* to the polymer template and find the most and least common elements in the result. *What do you get if you take the quantity of the most common element and subtract the quantity of the least common element?*

```{r input}
test <- read_lines("NNCB\n\nCH -> B\nHH -> N\nCB -> H\nNH -> C\nHB -> C\nHC -> B\nHN -> C\nNN -> C\nBH -> H\nNC -> B\nNB -> B\nBN -> B\nBB -> N\nBC -> B\nCC -> N\nCN -> C\n")

input <- read_lines("input_day14.txt")

parse_input <- function(.x) {
  
  .i <- str_detect(.x, "->")
  
  start <- .x[!.i & .x != ""]
  
  rules <- 
    tibble(x = .x[.i]) %>% 
    separate(x, c("pair", "insert"), " -> ")
  
  pairs <- start %>% 
    str_split("") %>% 
    pluck(1) %>% 
    str_c(lag(.), .) %>% 
    .[!is.na(.)] %>% 
    tibble(pair = .) %>% 
    count(pair)
  
  counts <- 
    start %>% 
    str_split("") %>% 
    pluck(1) %>% 
    tibble(chr = .) %>% 
    count(chr)
  
  .l <- list(pairs = pairs, counts = counts)
  
  assign("tbl_list", .l, globalenv())
  assign("rules", rules, globalenv())

}

```

```{r puzzle 1}
puz1 <- input

parse_input(puz1)

step <- function(.list, .ignore, .rules = rules) {
  
  counts <- .list$counts
  pairs <- .list$pairs 
  
  # for each step, first add new characters we're inserting to the "counts" table
  counts <- 
    pairs %>% 
    left_join(.rules, by = "pair") %>% 
    select(chr = insert, n) %>% 
    bind_rows(counts) %>% 
    count(chr, wt = n)
  
  # next, convert pairs we have to new pairs based on insert rules
  pairs <- 
    pairs %>% 
    left_join(.rules, by = "pair") %>% 
    mutate(first = str_c(str_sub(pair, end = 1), insert), 
           second = str_c(insert, str_sub(pair, start = 2)), 
           both = str_c(first, second, sep = ",")) %>% 
    select(both, n) %>% 
    separate_rows(both) %>% 
    rename(pair = both) %>% 
    count(pair, wt = n)
  
  list(pairs = pairs, counts = counts)
}

answer1 <- reduce(1:10, step, .init = tbl_list)

max(answer1$counts$n) - min(answer1$counts$n)

```

After 10 steps, from our input, when we take the quantity of the most common element and subtract the quantity of the least common element we get `2891`



# Puzzle 2 

The resulting polymer isn't nearly strong enough to reinforce the submarine. You'll need to run more steps of the pair insertion process; *a total of 40 steps should do it*.

In the above example, the most common element is B (occurring 2192039569602 times) and the least common element is H (occurring 3849876073 times); subtracting these produces 2188189693529.

*Apply 40 steps of pair insertion* to the polymer template and find the most and least common elements in the result. *What do you get if you take the quantity of the most common element and subtract the quantity of the least common element?*


```{r puzzle 2}
answer2 <- reduce(1:40, step, .init = tbl_list)

max(answer2$counts$n) - min(answer2$counts$n)
```

After 40 steps, the difference between our most common and least common letter element is `4607749009683`
